---
title: "Flight Delay Prediction"
author: "Devadharshini Udayakumar"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

# Abstract

This project analyzes the 2019 Airline Delays dataset to identify patterns contributing to flight delays. We focus on Frontier Airlines Inc. and use LDA and QDA models to predict delay status using weather-related variables.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(ggplot2)
library(readr)
library(MASS)
library(caret)
#library(GGally)
library(ggcorrplot)

```

# Load Data
```{r}
flights_data <- read.csv("/Users/devadharshini/Downloads/full_data_flightdelay.csv")
head(flights_data)
```

#Sample Selection

```{r}
library(dplyr)
airline_delay <- flights_data %>%
  filter(CARRIER_NAME == "Southwest Airlines Co.")  

airline_sample <- airline_delay %>%
  sample_n(1000)
#write_csv(airline_sample, "/Users/devadharshini/Desktop/5303-docs/Final_project/delta_sample_unbalanced.csv") 
```

## Data preprocessing 

```{r}
colnames(airline_sample)
```
```{r}
# Create DelayStatus column
library(dplyr)
airline_sample <- airline_sample %>%
  filter(!is.na(DEP_DEL15)) %>%
  mutate(DelayStatus = ifelse(DEP_DEL15 == 1, "Delayed", "OnTime"))



# Select relevant features
airline_sample <- airline_sample %>%
  select( MONTH, DAY_OF_WEEK, DISTANCE_GROUP,
         SEGMENT_NUMBER,CONCURRENT_FLIGHTS,AIRPORT_FLIGHTS_MONTH,
         AVG_MONTHLY_PASS_AIRPORT,
         AIRPORT_FLIGHTS_MONTH,
         PLANE_AGE, PRCP, SNOW, SNWD, TMAX, AWND,DelayStatus)

#Check for missing values
colSums(is.na(airline_sample)) 

```
There are no missing values.

## Exploratory Data Analysis

```{r}
library(ggplot2)
ggplot(airline_sample, aes(x = DelayStatus, fill = DelayStatus)) +
  geom_bar() +
  labs(title = "Flight Delay Status Distribution", x = "Status", y = "Count") +
  theme_minimal()
```


```{r}
summary(dplyr::select(airline_sample,PLANE_AGE, PRCP, SNOW, TMAX, AWND))
```
```{r}
library(tidyverse)

# Step 1: Reshape the data to long format
boxplot_data <- airline_sample %>%
  select(DelayStatus, PLANE_AGE, PRCP, SNOW, TMAX, AWND) %>%
  pivot_longer(cols = -DelayStatus, names_to = "Variable", values_to = "Value")

# Step 2: Create the combined boxplot with facets
ggplot(boxplot_data, aes(x = DelayStatus, y = Value, fill = DelayStatus)) +
  geom_boxplot() +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(
    title = "Boxplot of DelayStatus vs Flight and Weather Features",
    x = "Delay Status",
    y = "Value"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(face = "bold")
  )
#ggsave("/Users/devadharshini/Desktop/5303-docs/Final_project/facet_boxplots.png", width = 12, height = 8, dpi = 300)
```
Correlation Analysis:
```{r}
library(reshape2)
library(ggplot2)

# Step 1: Select only numeric columns
num_cols <- airline_sample[sapply(airline_sample, is.numeric)]

# Step 2: Compute correlation matrix
cor_matrix <- round(cor(num_cols, use = "complete.obs"), 2)

# Step 3: Mask the upper triangle
cor_matrix[upper.tri(cor_matrix)] <- NA

# Step 4: Melt to long format (only lower triangle remains)
melted_cor <- melt(cor_matrix, na.rm = TRUE)

# Step 5: Plot with ggplot2
ggplot(data = melted_cor, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = value), color = "black", size = 3) +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Correlation") +
  theme_minimal(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 9),
    axis.text.y = element_text(size = 9),
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  ) +
  labs(title = "Lower Triangular Correlation Matrix", x = NULL, y = NULL)

```
```{r}
cor_data <- dplyr::select(airline_sample, PRCP, SNOW, TMAX, AWND, PLANE_AGE)
cor_matrix <- cor(cor_data)

ggcorrplot(cor_matrix, lab = TRUE, colors = c("blue", "white", "red"),
           title = "Correlation Matrix of Weather Features")
```

# Modeling 
## Train-Test Split
```{r}
set.seed(42)

# Stratified split
train_index <- createDataPartition(airline_sample$DelayStatus, p = 0.7, list = FALSE)
train_data <- airline_sample[train_index, ]
test_data <- airline_sample[-train_index, ]

# Check class counts
table(train_data$DelayStatus)
table(test_data$DelayStatus)


```

## LDA Model 
```{r}
#train_data$DelayStatus <- factor(train_data$DelayStatus, levels = c("Delayed", "OnTime"))
#test_data$DelayStatus <- factor(test_data$DelayStatus, levels = c("Delayed", "OnTime"))

# Fit LDA model
lda_model <- lda(DelayStatus ~ ., data = train_data)
lda_pred <- predict(lda_model, test_data)

# Format predictions and actuals for confusionMatrix
predicted <- factor(lda_pred$class, levels = c("Delayed", "OnTime"))
actual <- factor(test_data$DelayStatus, levels = c("Delayed", "OnTime"))

confusionMatrix(predicted, actual)
```



## QDA Model 
```{r}
# Ensure target variables are consistent factors


# Train QDA model
qda_model <- qda(DelayStatus ~ . , data = train_data)

# Predict on test set
qda_pred <- predict(qda_model, test_data)

# Make sure predictions are treated as a factor with the same levels
predicted_qda <- factor(qda_pred$class, levels = c("Delayed", "OnTime"))
actual_qda <- factor(test_data$DelayStatus, levels = c("Delayed", "OnTime"))

# Confusion matrix
confusionMatrix(predicted_qda, actual_qda)
```


## Accuracy of LDA and QDA Comparison 
```{r}
# Calculate accuracy for both models
lda_acc <- mean(predicted == actual)
qda_acc <- mean(predicted_qda == actual_qda)

# Print comparison
accuracy_df <- data.frame(
  Model = c("LDA", "QDA"),
  Accuracy = c(round(lda_acc, 4), round(qda_acc, 4))
)

accuracy_df
```
The accuracy comparison shows that LDA slightly outperformed QDA, achieving 52.67% compared to QDA’s 49.67%. Although both are close to the 50% threshold (equivalent to guessing), LDA shows marginal improvement and may generalize slightly better for this classification task.

ROC Curve:
```{r roc-curve, fig.width=7, fig.height=5}
library(ROCR)

# Get LDA probabilities
lda_probs <- lda_pred$posterior[, "Delayed"]
pred_lda <- prediction(lda_probs, actual)
perf_lda <- performance(pred_lda, "tpr", "fpr")

# Get QDA probabilities
qda_probs <- qda_pred$posterior[, "Delayed"]
pred_qda <- prediction(qda_probs, actual_qda)
perf_qda <- performance(pred_qda, "tpr", "fpr")

# Plot ROC curves
plot(perf_lda, col = "blue", lwd = 2, main = "ROC Curve: LDA vs QDA")
plot(perf_qda, col = "red", lwd = 2, add = TRUE)
legend("bottomright", legend = c("LDA", "QDA"), col = c("blue", "red"), lwd = 2)
```
The ROC curve shows both LDA and QDA models performing close to the diagonal line, which suggests a low level of true signal distinguishing delayed from on-time flights. While LDA (blue) slightly outperforms QDA (red) in certain regions, both models hover around random guessing (AUC is approximate 0.5), indicating weak discriminative power when using weather features alone.


PCA + Decision Boundary

```{r}

library(MASS)
library(ggplot2)
library(dplyr)

# Step 1: Train LDA and QDA
lda_class <- predict(lda(DelayStatus ~ ., data = train_data), test_data)$class
qda_class <- predict(qda(DelayStatus ~ ., data = train_data), test_data)$class

# Step 2: PCA transformation on test data
pca_model <- prcomp(train_data %>% select(-DelayStatus), center = TRUE, scale. = TRUE)
pca_test <- predict(pca_model, test_data %>% select(-DelayStatus))

# Step 3: Create long-form data with model labels
pca_combined <- bind_rows(
  data.frame(PC1 = pca_test[, 1], PC2 = pca_test[, 2],
             Actual = test_data$DelayStatus,
             Predicted = lda_class, Model = "LDA"),
  
  data.frame(PC1 = pca_test[, 1], PC2 = pca_test[, 2],
             Actual = test_data$DelayStatus,
             Predicted = qda_class, Model = "QDA")
)

# Filter out LDA and QDA plots
pca_lda <- filter(pca_combined, Model == "LDA")
pca_qda <- filter(pca_combined, Model == "QDA")

# Plot LDA
p1 <- ggplot(pca_lda, aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Actual, fill = Predicted),
             color = "black", size = 3, stroke = 0.8, alpha = 0.75) +
  scale_shape_manual(values = c("Delayed" = 21, "OnTime" = 24)) +
  scale_fill_manual(values = c("Delayed" = "red", "OnTime" = "skyblue")) +
  labs(title = "LDA Prediction in PCA Space", subtitle = "Blue = Correct Prediction | Red = Wrong Prediction") +
  theme_minimal(base_size = 14)

# Plot QDA
p2 <- ggplot(pca_qda, aes(x = PC1, y = PC2)) +
  geom_point(aes(shape = Actual, fill = Predicted),
             color = "black", size = 3, stroke = 0.8, alpha = 0.75) +
  scale_shape_manual(values = c("Delayed" = 21, "OnTime" = 24)) +
  scale_fill_manual(values = c("Delayed" = "red", "OnTime" = "skyblue")) +
  labs(title = "QDA Prediction in PCA Space", subtitle = "Blue = Correct Prediction | Red = Wrong Prediction") +
  theme_minimal(base_size = 14)

# View plots separately
print(p1)
print(p2)
```

```
